---
title: "README"
author: "Jarrett Byrnes"
date: "December 10, 2014"
output: md_document
---
## A library to examine change in Sea Surface Temperatures

This library was developed following my own interest in replicating some of the analyses from Burrows et al. 2011 for SST change and climate velocity using the 
Met Office Hadley Centre Sea Surface Temperature 1.1 dataset, available at http://www.metoffice.gov.uk/hadobs/hadisst/data/download.html in netcdf format (the file HadISST1.1_sst_1870on.pp.gz gunzipped).

It provides a simple function to load the netcdf data.
```{r, cache=TRUE}
library(hadsstR)
sstData <- loadHadSST(directory="../", hadsstFilename="HadISST_sst.nc") 
summary(sstData)
```

One can then work with the dataset in a number of ways. For now, the functions all begin by creating annual averages, and working from there. The getClimateChange function generates a list of different matrices that can be used for further analyses of climate change.

```{r, cache=TRUE}
climateChangeMats <- getClimateChange(sstData, years=1960:2009)
```

We can plot a number of derived quantities from this object. For example, average temperature over the timespan
```{r}
#Let's plot some of these matrices
pal <- colorRampPalette(c("blue","white", "red"))

with(climateChangeMats, image(lon, lat, averageMat, col=pal(80)))

```

Or the slope of the regression between temperature and year for each lat/long cell
```{r}
library(lattice)
latLonGrid <- expand.grid(lon = climateChangeMats$lon, lat = climateChangeMats$lat)

levelplot(climateChangeMats$linearChangeMat ~ lon * lat, 
  data = latLonGrid, col.regions = pal(101), at=seq(-1,1,length.out=101))
```

We can look at spatial gradients in temperature
```{r}
pal2 <- colorRampPalette(c("green", "lightblue", "white", "yellow", "orange"))
with(climateChangeMats, image(lon, lat, spatialGradMat, col=pal2(101)))

levelplot(climateChangeMats$NSmat ~ lon * lat, col.regions=pal(100),
          data = latLonGrid, at=seq(-0.015, 0.015, length.out=100))

levelplot(climateChangeMats$EWmat ~ lon * lat, col.regions=pal2(100),
          data = latLonGrid, at=seq(-0.025, 0.025, length.out=100))
``` 

And with all of this, we can see climate change velocity
```{r}
#create a velocity matrix where values >200 and < -200 are truncated to those limits
#for easier plotting, as in Burrows et al. 20011
velMatTruncated <- climateChangeMats$velocityMat
velMatTruncated[velMatTruncated >200] <- 200
velMatTruncated[velMatTruncated < -200] <- -200

levelplot(velMatTruncated ~ lon * lat, data = latLonGrid, #at = cutpts, 
           pretty = T, 
          col.regions = pal(100),
           at=seq(-200,200,length.out=100))
```

Note that in the future I hope to add more functionality and deal with seasonal data. These methods *should* also work for other Hadley Centre data sets, but I have not yet tested them.


## Version numbers

`hadsstR` uses [semantic versioning][semver]. The version numbering
scheme is `major`.`minor`.`revision`. Unless `major` is 1, the package
should not be considered stable. All releases with the same `major`
versions are compatible. Increases in `minor` represents the addition of
backwards-compatible additions. Increases in `revision` represents either
bug fixes or improvements.


## Contributions

People wanting to contribute are welcome to do so by forking the repository,
and submitting a pull request when their work is done. Please also edit the
`DESCRIPTION` file to add your name to the `Authors` field.

[semver]: http://semver.org/